Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-12-03 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision based on Reto's feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2909 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0cc0d39aebda2df31d14542bffdc40f3d968a4f8

\Abstract{
Despite the occurrence of circular data in a wide range of scientific
fields, the existing methodology for probabilistically forecasting
circular data
is highly limited. Most of the existing methods are built on the
framework of
generalized linear and additive models, which can incorporate
difficulties in
both the optimization and the interpretation. As a very intuitive
alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models
simplify
the estimation process by separating the data into more homogeneous
subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each terminal node.  The proposed circular regression trees are easy to
interpret, can detect non-additive effects, and select covariates and their
interactions automatically. By combining an ensemble of circular regression
trees a more stable and regularized circular regression forest can be built
which allows for modeling smooth effects as well.

To evaluate the proposed circular regression trees and forests, a case
study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common
approaches
are used as a benchmark in order to assess the predictive performances of
circular regression trees and forests.
%Establishing a proper regression model for circular data such as animal movement
%directions, periodic measurements of neuronal activity or wind directions,
%requires specific attention regarding its representation on the unit circle.
%Commonly applied generalized linear and additive models can provide a useful toolbox, 
%however, highly non-linear link functions can complicate interpretations while
%a non-concave log-likelihood function of the usually employed von Mises distribution 
%leads to difficulties in the incorporated optimization procedure.
%Moreover, an adequate model specification can be very challenging and non-additive 
%effects might not be detected.

@@

In many approaches to model circular data it is assumed that the circular

@@

parameter of the von Mises distribution is predicted by a linear combination of

@@

Therefore, highly informative starting values are crucial for these circular GLMs to converge at all 

@@

due to the highly non-linear link function and the representation of smooth transitions 
on the unit circle is not straightforward. For example, the same rotation can be obtained in either 

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
only for linear but also for a circular covariate, the tree only considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
yielding a full probabilistic model. In addition, as a natural further development, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed to probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes especially in air traffic management.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a deep valley in the Alps.
Topography channels wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements at the airport (lagged by one hour), such as wind direction and
wind speed at the airport.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates employed are
  wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5). (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8). (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be distinguished clearly into wind regimes
and consequently are associated with a very low estimated concentration $\hat
\kappa$. In terms of covariates, the lagged wind direction (``persistence'') is
mostly responsible for distinguishing the broad wind regimes listed above while
the pressure gradients and wind speed separate between subgroups with high vs.\
low precision. Note that in the meteorological context wind direction 
is defined on the scale $[0,360]$ degree and increases clockwise from North ($0$ degree).

A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches. The
remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular regression trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and commonly applied for probabilistic modeling of circular
data. Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in (-\pi, \pi]$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

The corresponding log-likelihood function is defined by
%%
\begin{equation}
  \begin{aligned}
    \ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
    &= -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
  \end{aligned}
\end{equation}
%%

Once an appropriate distribution family is selected fitting a probabilistic
model corresponds to estimating the distribution parameters. Thus, when
applying the von Mises distribution to fit a probabilistic model $\text{vM}(Y;
\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ the
distribution parameters $\mu$ and $\kappa$ are to be estimated. For a learning
sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim} (\hat{\mu},\hat{\kappa}) =
\argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) \end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fit to the learning data.

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation and fitted parameter.
Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

@@

An alternative to additive models, which can deal or even 
avoid the mentioned difficulties in circular regression modeling, is to employ tree-based models. 
For the particular case of circular data, the methodology of circular regression trees and forests is
presented in the following section.

@@

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2. of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function $s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters and a data set of $n$ observations, this yields an
$n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).
A more detailed description of the applied tree-building algorithm
can be found in Appendix~\ref{app:algorithm}.

Once a distributional tree model is fitted it can be employed to obtain
probabilistic predictions for a possibly new set of observed covariates $\bm{z}
= (z_1, \ldots, z_m)$. Starting at the root node the tree structure leads the
observation to a terminal node, depending on the values of the covariates,
where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
corresponding subset of learning observations. This can also be expressed by
employing weights which indicate whether the $i$-th learning observation and
the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
out of $B$ segments partitioning the covariate space in disjoint subsets. In
that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
specifying the predicted von Mises distribution for a given $\bm{z}$ is
obtained by 
%%
\begin{equation} 
  (\hat{\mu},\hat{\kappa})(\bm{z}) =
  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
  \ell(\mu,\kappa; y_i).
\end{equation}
%%
Therefore, the same parameter pair is estimated for all observations belonging
to the same terminal node. This facilitates the application as the parameter
estimates do not need to be recalculated for each (new) observation via maximum
likelihood but can be extracted directly from the learning sample and the
fitted model.

While tree models can capture non-additive effects their structure and the
consequential strict separation of data into subgroups hinders an adequate
depiction of smooth effects. Therefore, they might impose abrupt changes even
in case of rather smooth transitions. This can be avoided by combining an
ensemble of trees in order to obtain a more stable forest model.

\subsection{Circular regression forests}
\label{sec:circforest} 
A natural extension of circular regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

@@

    location is often quite unstable and shows only a daily or annual seasonality

@@

    for the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model

@@

    For simplicity, rather to fit one single climatological model, we perform the
    model fitting separately rolling over all time points. For each time point, the
    training data set consists of the same daytime as the time of interest within
    31~days centered around the same calender day over all years except the year of
    interest. To gain full probabilistic climatological forecasts, we employ the

>>>>

\Abstract{
Despite the occurrence of circular data in a wide range of scientific
fields, the existing methodology for probabilistically forecasting
circular data
is highly limited. Most of the existing methods are built on the
framework of
generalized linear and additive models, which can incorporate
difficulties in
both the optimization and the interpretation. As a very intuitive
alternative
to additive circular regression models, we extend previous ideas of circular
trees and introduce probabilistic circular regression trees and forests
employing the von Mises distribution. The resulting tree-based models
simplify
the estimation process by separating the data into more homogeneous
subgroups
and fitting a simple intercept-only model for each resulting subset. By
incorporating the framework of distributional modeling, splits in all
distribution parameters are considered and a fully specified distribution is
provided for each terminal node.  The proposed circular regression trees are easy to
interpret, can detect non-additive effects, and select covariates and their
interactions automatically. By combining an ensemble of circular regression
trees a more stable and regularized circular regression forest can be built
which also allows for modeling smooth effects.

To evaluate the proposed circular regression trees and forests, a case
study on
probabilistic forecasting of wind direction for two airports in Austria is
presented. In addition to the tree-based models, alternative common
approaches
are used as a benchmark in order to assess the predictive performances of
the novel methodology of circular regression trees and forests.
%Establishing a proper regression model for circular data such as animal movement
%directions, periodic measurements of neuronal activity or wind directions,
%requires specific attention regarding its representation on the unit circle.
%Commonly applied generalized linear and additive models can provide a useful toolbox, 
%however, highly non-linear link functions can complicate interpretations while
%a non-concave log-likelihood function of the usually employed von Mises distribution 
%leads to difficulties in the incorporated optimization procedure.
%Moreover, an adequate model specification can be very challenging and non-additive 
%effects might not be detected.

@@

Many approaches to model circular data assume that the circular

@@

parameter of the von Mises distribution, is predicted by a linear combination of

@@

Therefore, highly informative starting values are crucial for these circular GLMs to converge 

@@

due to the highly non-linear link function and the not-straightforward representation of smooth 
transitions on the unit circle. For example, the same rotation can be obtained in either 

@@

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link functions must be employed in the estimation
of the group-specific distribution parameters which also simplifies the optimization process.
By incorporating the framework of distributional modeling, we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of statistical inference.
In addition, the employed tree structure allows to capture non-additive effects while forests 
enable the modeling of smooth changes.
Furthermore, covariates and their possible interactions do not need to be specified in advance as 
they are selected automatically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and further extensions \citep{Larraondo+Inza+Lozano:2018} do not only
allow for linear, but also for circular covariates, the tree solely considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
providing a full probabilistic model. In addition, as a natural extension, ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

@@

\subsubsection*{Motivating example}
To provide a first impression of the presented methodology a circular regression 
tree is employed for probabilistic wind direction forecasting. Wind direction is a 
classical circular quantity and accurate forecasts are of great importance for
decision-making processes, e.g., in air traffic management as considered in this study.
Figure~\ref{fig:tree_ibk} shows an estimated tree for 1-hourly forecasts at the
airport of Innsbruck which lies at the bottom of a narro valley within the European Alps.
Topography channels the wind along the west-east valley axis or along a tributary
valley intersecting from the south. Hence, pressure gradients to which valley
wind regimes react are considered as covariates along with other meteorological
measurements (lagged by one hour) and their derivatives, such as 
wind direction and wind speed at the airport or spatial and temporal differences.
%%
\begin{figure}[p!]
  \centering
  \includegraphics[width=0.95\textheight,,angle=90,origin=c]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
  \caption{Fitted tree based on the von Mises distribution for 1-hourly wind
  direction forecasts at the airport of Innsbruck. In each terminal node the
  empirical histogram (gray) and fitted density (red line) are depicted along
  with the estimated location parameter (red hand). The covariates used for splitting
  are wind direction (degree), wind speed ($\text{ms}^{-1}$), and pressure gradients
  ($\text{dpressure; hPa}$) west, east and south of the airport, all lagged by
  one hour.
  Note that in the meteorological context wind direction is defined on the 
  scale $[0,360]$ degree and increases clockwise from North ($0$ degree).
}
  \label{fig:tree_ibk} 
\end{figure}
%%
Together with the tree structure itself, Figure~\ref{fig:tree_ibk} depicts the
empirical (gray) and fitted von Mises (red) distribution of wind direction in
each terminal node. Based on the fitted location parameters $\hat \mu$, the
subgroups can be distinguished into the following wind regimes: (1)~Up-valley
winds blowing from the valley mouth towards the upper valley (from east to
west, nodes~4 and 5); (2) Downslope winds blowing across the Alpine crest
along the intersecting valley towards Innsbruck (from south-east to north-west,
node~8); (3) Down-valley winds blowing in the direction of the valley mouth
(from west to east, nodes 10, 12 and 13). Node~7 captures observations with
rather low wind speeds that cannot be clearly distinguished into specific wind 
regimes and are consequently associated with a very low estimated concentration 
parameter~$\kappa$, i.e. a high estimated variance. 
In terms of covariates, the lagged wind direction (``persistence'') is mostly 
responsible for distinguishing the broad range of wind regimes listed above while
the pressure gradients and wind speed separate the data into subgroups with high 
vs.\ low precision. 
A more extensive case study of circular regression trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.

The remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular regression trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

% -------------------------------------------------------------------
% Probabilistic circular modeling
% -------------------------------------------------------------------
%\newpage
\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper
circular probability distribution, i.e., a probability distribution which
accounts for the periodicity of circular data. Generally, this feature can be
obtained by ``wrapping'' the probability density function of any continuous
distribution around the unit circle \citep{Mardia+Jupp:2009}. In that way, the
wrapped Cauchy distribution or the wrapped normal distribution can be employed
to model symmetric unimodal circular data. A close approximation to the wrapped
normal distribution that is mathematically simpler and hence easier to use
\citep{Fisher:1993} is provided by the von Mises distribution, a purely
circular distribution which is also known as ``the circular normal
distribution'' and is a common choice for probabilistic modeling of circular
data. Based on a location parameter $\mu \in [0, 2\pi)$ and a concentration
parameter $\kappa > 0$ the density of the von Mises distribution for an
observation $y \in [0, 2\pi)$ is given by:
%%
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
%%
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order
$0$ \citep[see, e.g.,][for a more detailed
overview]{Jammalamadaka+Sengupta:2001}.

The corresponding log-likelihood function is defined as
%%
\begin{equation}
  \begin{aligned}
    \ell(\mu, \kappa; y) &= \log(f_\mathrm{vM}(y;\mu, \kappa)) \\
    &= -\log(2 \pi I_0(\kappa)) + \kappa \cos(y - \mu).\label{equ:vM:loglik}
  \end{aligned}
\end{equation}
%%

Once an appropriate distribution family is selected, fitting a probabilistic
model corresponds to estimating the distribution parameters. Thus, when
applying the von Mises distribution to fit a probabilistic model $\text{vM}(Y;
\mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ the
distribution parameters $\mu$ and $\kappa$ are to be estimated. For a learning
sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim} (\hat{\mu},\hat{\kappa}) =
\argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i) \end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that
a fully specified distributional model is fitted to the learning data.

By means of the score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
an indicator of deviation and in that way a measure of goodness of fit of the
model can be obtained for each observation and fitted parameter.
Moreover, the optimization problem in Equation~\ref{eq:optim} can also be
defined by 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
%%
\begin{figure}[t]
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \vspace{-1em}
    \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
  \begin{minipage}{.45\textwidth}
    \includegraphics[width=1.2\linewidth]{density_circular.pdf}
  \end{minipage}
  \begin{minipage}{.05\textwidth}
    \hfill
  \end{minipage}%
  \caption{Illustration of a distributional model fitted to circular data by
    maximum likelihood estimation employing the von Mises distribution. Left
    panel: linearized illustration restricted to the interval $[0, 2\pi)$. Right
    panel: circular illustration. In both panels the empirical histogram (gray)
    and fitted density (red line) are depicted along with the estimated location
    parameter (red hand).}
  \label{fig:densities} 
\end{figure}
%%

@@

An alternative to additive models, which can deal or even 
avoid the mentioned difficulties in circular regression modeling, are 
tree-based models. For the particular case of circular data, the methodology 
of circular regression trees and forests is presented in the following section.

@@

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with strong variations. Therefore, separating the data set
into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (possibly) group specific
effects more precisely and hence can result in an overall better fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the
unbiased recursive partitioning algorithms MOB
\citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. By applying the presented distributional
trees, a full distributional model is specified in each node of the tree for a
specific distribution family chosen in advance. Hence, selecting the von Mises
distribution allows for an application of distributional trees to circular data
following the tree building algorithm as introduced in Section~2.2 of
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on
how and where to split the data is based on model scores which are obtained by
evaluating the score function~$s$ for the individual observations and parameter
estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its
two distribution parameters ($\mu$ and $\kappa$) and a data set of $n$ observations, 
this yields an $n \times 2$ matrix that can be employed as a discrepancy measure, capturing
how well each given observation conforms with the estimated location
$\hat{\mu}$ and precision $\hat{\kappa}$, respectively. To capture dependence
on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a
permutation test~(CTree). By doing so in each partitioning step, the covariate
with the highest significant association (i.e., lowest significant $p$-value,
if any) is selected for splitting the data. The corresponding split point is
chosen either by optimizing the log-likelihood~(MOB) or a two-sample test
statistic~(CTree) over all possible partitions. This procedure is repeated
recursively until there are no significant parameter instabilities or until
another stopping criterion is met (e.g., subgroup size or tree depth).
A more detailed description of the applied tree-building algorithm
can be found in Appendix~\ref{app:algorithm}.

Once a distributional tree model is fitted it can be employed to obtain
probabilistic predictions for a possibly new set of observed covariates $\bm{z}
= (z_1, \ldots, z_m)$. Starting at the root node, the tree structure leads the
observation to a terminal node, depending on the values of the covariates,
where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for the
corresponding subset of learning observations. This can also be expressed by
employing weights which indicate whether the $i$-th learning observation and
the observation $\bm{z}$ belong to the same terminal node: \begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in
\mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)), \end{equation} where
$\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th
out of $B$ segments partitioning the covariate space in disjoint subsets. In
that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$
specifying the predicted von Mises distribution for a given $\bm{z}$ is
obtained by 
%%
\begin{equation} 
  (\hat{\mu},\hat{\kappa})(\bm{z}) =
  \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot
  \ell(\mu,\kappa; y_i).
\end{equation}
%%
Therefore, the same parameter pair is estimated for all observations belonging
to the same terminal node. This facilitates the application as the parameter
estimates do not need to be recalculated for each (new) observation via maximum
likelihood but can be extracted directly from the learning sample and the
fitted model.

While tree models can capture non-additive effects, their structure and the
consequential strict separation of data into subgroups hinders an adequate
depiction of smooth effects. Therefore, they might impose abrupt changes even
in case of rather smooth transitions. This can be avoided by combining an
ensemble of trees in order to obtain a more stable forest model.

\subsection{Circular regression forests}
\label{sec:circforest} 
A natural extension of (circular) regression trees are ensembles or forests
that can improve forecasts by regularizing and stabilizing the model. Random
forests introduced by \cite{Breiman:2001} average the predictions of an
ensemble of trees, each built on a subsample or bootstrap of the original data.
A generalization of this strategy is to obtain weighted predictions by adaptive
local likelihood estimation of the distributional parameters
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019,
Hothorn+Zeileis:2017}. More specifically, for each possibly new
observation~$\bm{z}$ a set of averaged ``nearest neighbor'' weights
$w_i^{forest}(\bm{z})$ is obtained that is based on the number of trees in
which $\bm{z}$ is assigned to the same terminal node as each learning
observation $y_i, i \in \{1,\ldots,n\}$. Hence, for a forest of $T$ trees, the
weights are calculated via
%%
\begin{equation}
  w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
  \frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
%%
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree. Therefore, similar observations ending up more
often in the same terminal node have higher weights and in that way a stronger
influence in the estimation process.

@@

    location is often quite unstable and only shows a daily cycle or an annual seasonality

@@

    in the analysis of Figure~\ref{fig:tree_ibk}. Hence, the climatological model

@@

    For simplicity, rather to fit one single climatological model, a single model is estimated 
    rolling over all observation time points, 
    using the $31$~days centered around the day of interest for all available years in the training data.
    %For simplicity, rather to fit one single climatological model, we perform the
    %model fitting separately rolling over all time points. For each time point, the
    %training data set consists of the same daytime as the time of interest within
    %31~days centered around the same calender day over all years except the year of
    %interest. 
    To gain full probabilistic climatological forecasts, we employ the
